\documentclass{report}
    \title{50001 - Algorithm Analysis and Design - Lecture 2}
    \author{Oliver Killane}
    \date{12/11/21}

\input{../50001 common.tex}

\newcommand{\cond}[3]{\text{if } #1 \text{ then } #2 \text{ else } #3}

\begin{document}
\maketitle
\lectlink{https://imperial.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=c8abc020-d7c8-44ad-b1df-adc100991181}

\section*{Evaluation \& Cost Models}
\codelist{Haskell}{minimum.hs}
When analysing the cost of \fun{minimum} we must consider how the function is evaluated.
\\
\\ For example we could shortcut the once \fun{isort} has determined the first element (the minimum) of the list.
\\
\sidenote{Cost Model}{
	A model to determine the time taken to execute a program.
	\\
	\\ The model assigns cost to different operations (e.g comparisons, calls, memory reads/writes)
	\\
	\\ A very generalised cost model assigns cost based on the number of \keyword{reductions} required to evaluate a program.
}

\section*{Small While Language}
We can define a small language of expressions as follows:
\[e ::= x \ | \ k \ | \ f \ e_1 \dots e_n \ | \ \cond{e}{e_1}{e_2}\]
where $k$ means constant and $x$ is the variable form.
\\ Infix functions such as $+, -, \times$ are written normally, and are also expressions as they can be used in the form $(+) \ e_1 \ e_2$.
\\
\\ There are also several primitive constants: $True, False, 0, 1, 2, \dots$
\\ List constants and operations are also primitive: $[], (:), null, head, tail$

\section*{Evaluation Order}
\begin{itemize}
	\bullpara{Applicative Order}{ Strict evaluation
		\\The leftmost, innermost reducible expression is evaluated first.
		\\ e.g for $fst (3 \times 2, 1 + 2)$
		\\ \begin{steps}
			\start{fst (3 \times 2, 1 + 2)}
			\step{ fst (6, 1 + 2)}{Definition of $\times$}
			\step{fst (6, 3)}{ Definition of $+$}
			\step{6}{Definition of $fst$}
		\end{steps}
	}
	\bullpara{Normal Order}{ Lazy evaluation
		\\ The leftmost outer reducible is evaluated first. Efectively evaluating the function before its arguments.
		\\ e.g for $fst (0, 1 + 2)$
		\\ \begin{steps}
			\start{fst (3 \times 2, 1 + 2)}
			\step{3 \times 2}{ Definition of $fst$}
			\step{6}{ Definition of $\times$}
		\end{steps}
	}
\end{itemize}
For a given program, if \keyword{applicative} and \keyword{normal} terminate, then they produce the same value in normal form.
\\
\\ However there are some programs where \keyword{normal} evaluation terminates, but \keyword{applicative} will not.
\begin{minipage}[t]{0.4\textwidth}
	\centerline{\textbf{Applicative}}
	\begin{steps}
		\start{fst (0, \text{crazy nonesense})}
		\step{CRASH!}{ By lack of definition for crazy nonesense}
	\end{steps}
\end{minipage}
\hfill
\begin{minipage}[t]{0.4\textwidth}
	\centerline{\textbf{Normal}}
	\begin{steps}
		\start{fst (0, \text{crazy nonesense})}
		\step{0}{ Definition of $fst$}
	\end{steps}
\end{minipage}
The program may by syntactically correct, but have an error such as zero-division which will not be evaluated and hence not result in improper termination under \keyword{normal} order.
\begin{large} \[\text{Applicative Terminates} \Rightarrow \text{Normal Terminates}\] \end{large}

\section*{Cost Model for Small While}
We can evaluate a cost model for the small while language by creating a function $T$ to assign cost to expressions.
\\\begin{tabular}{p{0.25\textwidth} p{0.35\textwidth} p{0.4\textwidth}}
	\textbf{Type}          & \textbf{Function}                                                               & \textbf{Explanation}                                                                                                                \\
	\hline
	non-primitive function & $\begin{matrix}
			f \ a_1 \ \dots \ a_n = e           \\
			T(f) \ a_1 \ \dots \ a_n = T(e) + 1 \\
		\end{matrix}$                                                     & Given we have already computed all argument, the cost of the function is the cost of the expression it produces, and a single call. \\
	\hline
	primitive function     & $T(f) \ x \dots \ x_n = 0$                                                      & Primitive functions are assumed to be free.                                                                                         \\
	\hline
	Variable               & $T(x) = 0$                                                                      & accessing variables is free.                                                                                                        \\
	\hline
	Application            & $T(f \ e_1 \ \dots \ e_n) = T(f) \ e_1 \ \dots \ e_n + T(e_1) + \dots + T(e_n)$ & When applying a function we must consider both its cost, and the cost of all argument expressions.                                  \\
	\hline
	Conditional            & $T(\cond{p}{e_1}{e_2}) = T(p) + \cond{p}{T(e_1)}{T(e_2)}$                       & Cost of condition and of the resulting expression.                                                                                  \\
\end{tabular}

\section*{Cost Model Example}
Given the function:
\[mul \ m \ n = \cond{m=0}{0}{n + mul \ (m-1) \ n}\]
Evaluate $T(mul \ 3 \ 100)$
\\
\begin{proof}
	\proofstep{1}{$mul \ 3 \ 100$}{}
	\proofstep{2}{$T(\cond{3=0}{0}{100 + mul \ (3-1) \ 100}) + 1$}{By Rule for non-primitive functions}
	\proofstep{3}{$T(3=0) + T(100 + mul \ (3-1) \ 100) + 1$}{By rule for conditionals}
	\proofstep{4}{$0 + T(100 + mul \ (3-1) \ 100) + 1$}{By primitive functions}
	\proofstep{5}{$T(+) (100 \ mul \ (3-1) \ 100) + T(100) + T(mul \ (3-1) \ 100) + 1$}{By application rule}
	\proofstep{6}{$0 + T(100) + T(mul \ (3-1) \ 100) + 1$}{By rule for primitive functions}
	\proofstep{7}{$0 + T(mul \ (3-1) \ 100) + 1$}{By rule for constants}
	\proofstep{8}{$T(mul) \ (3-1) \ 100 + T(3 - 1) + T(100) + 1$}{By application rule}
	\proofstep{9}{$T(mul) \ (3-1) \ 100 + T(-) 3 \ 1 + T(100) + 1$}{By application rule}
	\proofstep{10}{$T(mul) \ 2 \ 100 + 1$}{By application rule}
	\proofstep{11}{$T(\cond{2=0}{0}{100 + mul \ (2-1) \ 100}) + 1 + 1$}{By Rule for non-primitive functions}
	\proofstep{12}{$T(2=0) + T(100 + mul \ (2-1) \ 100) + 2$}{By rule for conditionals}
	\proofstep{13}{$0 + T(100 + mul \ (2-1) \ 100) + 2$}{By primitive functions}
	\proofstep{14}{$T(+) (100 \ mul \ (2-1) \ 100) + T(100) + T(mul \ (2-1) \ 100) + 2$}{By application rule}
	\proofstep{15}{$0 + T(100) + T(mul \ (2-1) \ 100) + 2$}{By rule for primitive functions}
	\proofstep{16}{$0 + T(mul \ (2-1) \ 100) + 2$}{By rule for constants}
	\proofstep{17}{$T(mul) \ (2-1) \ 100 + T(2 - 1) + T(100) + 2$}{By application rule}
	\proofstep{18}{$T(mul) \ (2-1) \ 100 + T(-) 2 \ 1 + T(100) + 2$}{By application rule}
	\proofstep{19}{$T(mul) \ 1 \ 100 + 2$}{By application rule}
	\proofstep{20}{$T(\cond{1=0}{0}{100 + mul \ (1-1) \ 100}) + 2 + 1$}{By Rule for non-primitive functions}
	\proofstep{21}{$T(2=0) + T(100 + mul \ (1-1) \ 100) + 3$}{By rule for conditionals}
	\proofstep{22}{$0 + T(100 + mul \ (1-1) \ 100) + 3$}{By primitive functions}
	\proofstep{23}{$T(+) (100 \ mul \ (1-1) \ 100) + T(100) + T(mul \ (1-1) \ 100) + 3$}{By application rule}
	\proofstep{24}{$0 + T(100) + T(mul \ (1-1) \ 100) + 3$}{By rule for primitive functions}
	\proofstep{25}{$0 + T(mul \ (1-1) \ 100) + 3$}{By rule for constants}
	\proofstep{26}{$T(mul) \ (1-1) \ 100 + T(1 - 1) + T(100) + 3$}{By application rule}
	\proofstep{27}{$T(mul) \ (1-1) \ 100 + T(-) 2 \ 1 + T(100) + 3$}{By application rule}
	\proofstep{28}{$T(mul) \ 0 \ 100 + 3$}{By application rule}
	\proofstep{29}{$T(\cond{0=0}{0}{100 + mul \ (1-1) \ 100}) + 3 + 1$}{By Rule for non-primitive functions}
	\proofstep{30}{$T(0=0) + T(0) + 4$}{By rule for conditionals}
	\proofstep{31}{$0 + T(0) + 4$}{By rule for primitive functions}
	\proofstep{32}{$0 + 4$}{By rule for variables}
	\proofstep{33}{$4$}{}
\end{proof}
\end{document}
